Grocery Pricing Strategies: An Analysis of Vendor Positioning and Market Dynamics

### 2024-11-14 paper/paper.qmd
update paper, llm, readme

# Data {#sec-data}
We use the statistical programming language R [@citeR] and several libraries including tidyverse [@tidyverse], DBI [@citeDBI], RSQLite [@citeRSQLite], and ggplot2 [@citeGG]. The data was sourced from a project database created by Jacob Filipp, containing grocery price data collected across various vendors.

@@

# Discussion {#sec-discussion}
The results of this analysis reveal distinct pricing and promotional strategies among the grocery vendors, indicating how each vendor may be positioning itself within the market. Metro’s high frequency of sales, shown in the sales frequency graph, suggests a competitive pricing strategy that relies heavily on promotional events to attract customers. This approach could be aimed at creating a perception of savings, encouraging frequent purchases, and catering to a broad range of customers who respond to discounts. The presence of frequent sales might also reflect Metro’s intention to balance a diverse range of price points, as seen in the wide distribution of prices across products. This flexibility allows Metro to cater to both budget-conscious consumers and those willing to pay premium prices.

On the other hand, No Frills and Walmart demonstrate a consistent commitment to low prices, with lower price distributions and the lowest average prices among the vendors. Their relatively low frequency of sales suggests a focus on everyday low pricing rather than relying on promotions to drive customer traffic. This approach is likely attractive to price-sensitive shoppers who value predictably low prices over the fluctuations associated with frequent sales. The data implies that No Frills and Walmart have established themselves as budget-friendly options, strategically positioning themselves as reliable choices for consumers seeking affordability.

Galleria, with a high average price and fewer sales, appears to target a different segment. The relatively high prices suggest a focus on specialty or premium products, positioning Galleria as a niche or premium vendor within the market. This positioning may attract consumers who prioritize quality or unique offerings over low prices. The lack of high-frequency sales events implies a strategy that focuses less on discount-driven customer acquisition and more on brand or product differentiation, possibly appealing to a specific audience that values specialty goods.

The remaining vendors, including Loblaws, Save-On-Foods, T&T, and Voila, demonstrate a more balanced approach. These vendors have moderate frequencies of sales and average prices, positioning them between the budget-oriented and premium segments. This strategy likely appeals to a broad consumer base that includes both cost-conscious shoppers and those willing to spend slightly more for a variety of options. By maintaining average prices that are neither the lowest nor the highest, these vendors may capture customers seeking value without the extremes of budget or premium positioning.

Overall, the data illustrates a competitive landscape where each vendor’s pricing and promotional strategy plays a distinct role in shaping customer perception and loyalty. Metro’s aggressive promotional tactics, combined with a wide range of prices, allow it to attract a diverse audience. Meanwhile, No Frills and Walmart stand out as dependable low-cost options, appealing to value-driven customers. Galleria’s premium positioning caters to a niche market segment, and the remaining vendors capture a middle ground, appealing to customers looking for a balance of price and quality. These findings underscore the varied strategies used by grocery vendors to differentiate themselves and compete for different consumer segments in a competitive market.

>>>>

This paper is organized as follows: @sec-data outlines the dataset and its measurement of real-world pricing strategies. @sec-results presents key findings on sales frequency, price distribution, and average price, revealing distinct vendor pricing strategies. @sec-discussion addresses considerations such as correlation versus causation, missing data, and sources of bias, which are essential for interpreting the analysis accurately.

# Data {#sec-data}
We use the statistical programming language R [@citeR] and several libraries, including tidyverse [@tidyverse], DBI [@citeDBI], RSQLite [@citeRSQLite], and ggplot2 [@citeGG]. The data was sourced from Project Hammer [@Filipp_2024], a project database created by Jacob Filipp that compiles grocery price data from various vendors.

## Measurement
The dataset used in this analysis captures real-world grocery pricing and promotional strategies from multiple vendors. Each entry, sourced from Project Hammer [@Filipp_2024], represents a product listing, with variables such as current_price, old_price, vendor, and nowtime. These fields translate vendor pricing strategies into data points: current_price reflects the active selling price, while old_price denotes any discount or promotional pricing. The vendor field identifies the grocery chain, and nowtime timestamps each entry, allowing us to analyze trends over time. This structure enables an organized examination of pricing tactics across vendors by converting complex market behaviors into measurable data.

@@

# Discussion {#sec-discussion}
The results of this analysis offer valuable insights into the pricing and promotional strategies of various grocery vendors. However, several methodological considerations need to be addressed to ensure an accurate interpretation of the data. Understanding the distinction between correlation and causation, handling missing data, and accounting for potential sources of bias are essential to refining the conclusions drawn from this study. Each of these factors is discussed in the following subsections.

## Correlation vs. Causation
In analyzing grocery pricing and promotional strategies, it’s important to distinguish between correlation and causation. While our data reveals correlations—such as a relationship between high-frequency sales and lower average prices in certain vendors—these associations do not imply causation. For example, Metro’s high sales frequency correlates with its broader price range, but this does not necessarily mean that frequent sales cause price diversity, or vice versa. Factors like target market segmentation, inventory management, and strategic pricing decisions can all influence these metrics independently. Therefore, while the analysis highlights patterns in how different vendors structure pricing and promotions, it does not establish causative relationships between these variables.

## Missing Data
The dataset may contain missing values, particularly in fields such as current and old prices. These missing values can affect the accuracy of the analysis, especially in sections where price distribution and averages are calculated. For example, if a vendor has missing values during high promotional periods, their frequency of sales may be underestimated. Similarly, missing prices for premium products could bias the average price calculation downward. To mitigate this, we addressed missing data by filtering out incomplete rows for critical analyses; however, the absence of data for certain time periods or products still limits the representativeness of our findings.

## Sources of Bias
Several potential sources of bias could affect the analysis and interpretation of grocery pricing data. First, the data is derived from a limited selection of vendors, which may not represent all grocery stores in the region, particularly smaller, independent stores. Second, the data is based on specific geographic regions and may not account for regional pricing variations. Finally, since the data was collected from website listings, it may not reflect in-store promotions or exclusive discounts for loyalty program members, thus potentially underestimating actual promotional frequencies. These sources of bias should be considered when interpreting the results, as they may influence conclusions about vendor strategies and market positioning.
